Modelling the emergence of whisker barrels

### 2020-04-21 paper/briefpaper/brief.tex

\begin{abstract}
Brain development relies on an interplay between genetic specification and
self-organisation. Striking examples of this relationship can be found in the
somatosensory brainstem, thalamus, and cortex of rats and mice, where the
arrangement of the facial whiskers is preserved in the arrangement of cell
aggregates to form precise somatotopic maps. We show in simulation how
realistic whisker maps can self-organize, by assuming that information is
exchanged between adjacent cells only, under the guidance of gene expression
gradients. The resulting model provides a simple account of how patterns of
gene expression can constrain spontaneous pattern formation to faithfully
reproduce functional maps in subsequent brain structures.
\end{abstract}

@@

The center-point and reaction-diffusion models are not mutually
exclusive. Pre-organised centers could bias reaction-diffusion processes to
generate specific arrangements more reliably, and mechanisms of lateral axon
branching may constitute the tension between cooperation and competition
required for self-organization. However, proof that barrel patterning can
emerge from an undifferentiated bundle of axons, based only on local
interactions, would show that a separate stage and/or extrinsic mechanism for
pre-organising thalamocortical connections need not be assumed. To this end,
we ask whether barrel maps can emerge in a system with reaction-diffusion
dynamics, under the guidance of signalling gradients, and in the absence of
pre-defined centers.

@@

First we verified that all results established by \cite{karbowski_model_2004}
for a 1D axis could be reproduced using our extension to a 2D cortical
sheet. Using an elliptical domain, $S$, with $M=3$ offset guidance gradients
aligned to the longer axis, $N=5$ thalamocortical projections gave rise to
five distinct cortical fields at locations that preserved the topographic
ordering defined by the original $\gamma$ values. However, we found that
specifying $N$ ordered areas required $M\approx (N+1)/2$ signalling
fields. This is because localization of axon densities occurs only when
projections are influenced by interactions with two or more signalling
gradients that encourage migration in opposing directions. Note that these
dynamics are quite unline classic chemospecificity models
\citep{sperry_chemoaffinity_1963}, which essentially assume center-points
i.e., conditions in the target tissue that instruct pre-identified afferents
to stop growing. As the number of distinct guidance fields is unlikely to
approach the number of barrels, further extensions were required.

The term in parentheses in Eq.\,\ref{eq:dc} represents competition between
thalamocortical projections for a limited availability of cortical
connections. To introduce competition also in terms of axon branching we
redefined
%
\begin{equation} \label{eq:comp}
\chi_i(\mb{x}, t) = - \frac{\epsilon  a_i}{N-1} \sum_{j \ne i}^{N} a_{j},
\end{equation}
%
which reduces branching for each projection where the branches of other
projections are dense. Divisive normalization keeps the axon branch density
bounded on each iteration to the initial (random) density of unconnected
  branches at $t=0$
%
\begin{equation} \label{eq:norm}
  a_i(\mathbf{x}, t) = \oiint_{S} \big[ a_i(\mathbf{x}, 0) - c_i(\mathbf{x}, t) \big] \dif S \; \frac {a'_i(\mathbf{x}, t)} {\oiint_{S}
  a'_i(\mathbf{x}, t) \dif S}\,,
\end{equation}
%
where prime symbols indicate the use of intermediate values computed from
Eqs.~\ref{eq:dc}--\ref{eq:comp} at time $t-{\delta}t$. Note that this
  operation is local to individual afferent projections.

The only differences between thalamic projections are their strengths of
interaction with the guidance fields, $\gamma$, hence any reliable differences
between emergent cortical fields must be due to differences in these values
only. We speculate that the contribution of a given ephrin field, to the
velocity at which a projection migrates across the cortical subplate, is
determined by the concentration of a similar molecule at its thalamic origin,
i.e., the putative barreloid center. As such, two orthogonal linear thalamic
gradients were defined, from which 41 pairs of $\gamma$ values were sampled,
at the coordinates of 41 barreloid centers recreated from
\citealp{haidarliu_size_2001} (Fig.\,1B).

A cortical boundary enclosing 41 corresponding barrels was traced from a
cytochrome oxidase stain from \cite{zheng_signal_2001}, and
Eqs.~\ref{eq:dc}--\ref{eq:norm} were solved for $N=41$ projections on the
resulting domain, using $M=2$ linear signalling gradients aligned with the
anterior-posterior and medial-lateral axes. From random initial conditions for
$a({\bf x},0)$ and $c({\bf x},0)$, a clear Voronoi-like tesselation emerged
(Fig.\,1C; see Movie S1). A reduction in the Honda delta metric (see
\citealp{senft_mouse_1991}) confirmed that a `good' Voronoi pattern emerged
within $\approx 3000$ iterations (see Fig.\,1D). A measure of the difference
in shape between real and simulated barrels revealed a strong correspondence
(see Fig.\,1D).

@@

The emergence of barrels in simulation required competition between thalamic
projections in terms of synaptic connectivity and an implicit self/other
identifier amoungst projections, and also competition in terms of cortical
space (as represented by $\chi$). This latter requirement may account for the
absence of
barrels in rodents with larger brains, such as capybara, for which competition
for space is presumably weaker \citep{woolsey_comparative_1975}. Hence,
irrespective of whether barrels are necessary for adaptive whisker function,
the emergence of somatotopically ordered modular structures may be an
inevitable consequence of local competition for cortical territory driven by
input from an array of discrete sensory organs \citep{purves_iterated_1992}.

It is important to emphasize that the formulation of the model is entirely
local, insofar as simulation requires no information to be communicated from a
given cortical grid cell to any but those immediately adjacent (via
diffusion). Hence the simulations demonstrate how a self-organising system,
constrained by genetically specified guidance cues and by the shape of the
cortical field boundary, can faithfully reproduce an arrangement of cell
aggregates in one neural structure as a topographic map in another.

>>>>

\begin{abstract}
Brain development relies on an interplay between genetic specification and
self-organization. Striking examples of this relationship can be found in the
somatosensory brainstem, thalamus, and cortex of rats and mice, where the
arrangement of the facial whiskers is preserved in the arrangement of cell
aggregates to form precise somatotopic maps. We show in simulation how
realistic whisker maps can self-organize, by assuming that information is
exchanged between adjacent cells only, under the guidance of gene expression
gradients. The resulting model provides a simple account of how patterns of
gene expression can constrain spontaneous pattern formation to faithfully
reproduce functional maps in subsequent brain structures.
\end{abstract}

@@

The center-point and reaction-diffusion models are not mutually
exclusive. Pre-organized centers could bias reaction-diffusion processes to
generate specific arrangements more reliably, and mechanisms of lateral axon
branching may constitute the tension between cooperation and competition
required for self-organization. However, proof that barrel patterning can
emerge from an undifferentiated bundle of axons, based only on local
interactions, would show that a separate stage and/or extrinsic mechanism for
pre-organizing thalamocortical connections need not be assumed. To this end,
we ask whether barrel maps can emerge in a system with reaction-diffusion
dynamics, under the guidance of signalling gradients, and in the absence of
pre-defined centers.

@@

First we verified that all results established by \cite{karbowski_model_2004}
for a 1D axis could be reproduced using our extension to a 2D cortical
sheet. Using an elliptical domain, $S$, with $M=3$ offset guidance gradients
aligned to the longer axis, $N=5$ thalamocortical projections gave rise to
five distinct cortical fields at locations that preserved the topographic
ordering defined by the original $\gamma$ values. However, we found that
specifying $N$ ordered areas required $M\approx (N+1)/2$ signalling
fields. This is because localization of axon densities occurs only when
projections are influenced by interactions with two or more signalling
gradients that encourage migration in opposing directions. Note that these
dynamics are quite unlike classic chemospecificity models
\citep{sperry_chemoaffinity_1963}, which essentially assume center-points
i.e., conditions in the target tissue that instruct pre-identified afferents
to stop growing. As the number of distinct guidance fields is unlikely to
approach the number of barrels, further extensions were required.

The term in parentheses in Eq.\,\ref{eq:dc} represents competition between
thalamocortical projections for a limited availability of cortical
connections. To introduce competition also in terms of axon branching we
redefined
%
\begin{equation} \label{eq:comp}
\chi_i(\mb{x}, t) = - \frac{\epsilon  a_i}{N-1} \sum_{j \ne i}^{N} a_{j},
\end{equation}
%
which reduces branching for each projection where the branches of other
projections are dense. Divisive normalization keeps the axon branch density
bounded on each iteration to the initial (random) density of unconnected
  branches at $t=0$
%
\begin{equation} \label{eq:norm}
  a_i(\mathbf{x}, t) = \oiint_{S} \big[ a_i(\mathbf{x}, 0) - c_i(\mathbf{x}, t) \big] \dif S \; \frac {a'_i(\mathbf{x}, t)} {\oiint_{S}
  a'_i(\mathbf{x}, t) \dif S}\,,
\end{equation}
%
where prime symbols indicate the use of intermediate values computed from
Eqs.~\ref{eq:dc}--\ref{eq:comp} at time $t-{\delta}t$ (${\delta}t$ is the time
represented by one simulation step). Note that this
  operation is local to individual afferent projections.

The only differences between thalamic projections are their strengths of
interaction with the guidance fields, $\gamma$, hence any reliable differences
between emergent cortical fields must be due to differences in these values
only. We speculate that the contribution of a given ephrin field, to the
velocity at which a projection migrates across the cortical subplate, is
determined by the concentration of a similar molecule at its thalamic origin,
i.e., the putative barreloid center. As such, two orthogonal linear thalamic
gradients were defined, from which 41 pairs of $\gamma$ values were sampled,
at the coordinates of 41 barreloid centers recreated from
\cite{haidarliu_size_2001} (Fig.\,1B).

A cortical boundary enclosing 41 corresponding barrels was traced from a
cytochrome oxidase stain from \cite{zheng_signal_2001}, and
Eqs.~\ref{eq:dc}--\ref{eq:norm} were solved for $N=41$ projections on the
resulting domain, using $M=2$ linear signalling gradients aligned with the
anterior-posterior and medial-lateral axes. From random initial conditions for
$a(\mb{x},0)$ and $c(\mb{x},0)$, a clear Voronoi-like tesselation emerged
(Fig.\,1C; see Movie S1). A reduction in the Honda delta metric (see
\citealp{senft_mouse_1991}) confirmed that a `good' Voronoi pattern emerged
within $\approx 3000$ iterations (Fig.\,1D). A measure of the difference
in shape between real and simulated barrels revealed a strong correspondence
(see Fig.\,1D).

@@

The emergence of barrels in simulation required competition between thalamic
projections in terms of synaptic connectivity and also competition in terms of
cortical space, as represented by $\chi$, with an implicit requirement for a
self/other identifier amoungst projections. This latter form of competition may
account for the absence of barrels in rodents with larger brains, such as
capybara, for which competition for space is presumably weaker
\citep{woolsey_comparative_1975}. Hence, irrespective of whether barrels are
necessary for adaptive whisker function, the emergence of somatotopically
ordered modular structures may be an inevitable consequence of local
competition for cortical territory driven by input from an array of discrete
sensory organs \citep{purves_iterated_1992}.

It is important to emphasize that the formulation of the model is entirely
local, insofar as simulation requires no information to be communicated from a
given cortical grid cell to any but those immediately adjacent (via
diffusion). Hence the simulations demonstrate how a self-organizing system,
constrained by genetically specified guidance cues and by the shape of the
cortical field boundary, can faithfully reproduce an arrangement of cell
aggregates in one neural structure as a topographic map in another.
